---
title: "161.324 Data Mining"
subtitle: "Assignment 1, 2026"
author: "Emma Bridge"
output: html_document
---

```{r setup, echo=TRUE, warning=FALSE, message=FALSE}
# Add any other packages you need to load here.
library(tidyverse)


# Read in the data
climate <- read_csv("https://www.massey.ac.nz/~jcmarsha/161324/assessment/data/climate.csv")
athletes <- read_csv("https://www.massey.ac.nz/~jcmarsha/161324/assessment/data/athletes.csv")
```

**THIS FILE IS WHERE YOU WRITE YOUR ANSWERS. THE QUESTIONS ARE AVAILABLE ON STREAM**

## Exercise 1: Analysis of Incomplete Temperature

### 1.1. Precise year and month

```{r}



```

### 1.2. Table of mean temperature by month for locations C and D

```{r}

```

### 1.3. Monthly temperatures at location C in 1919

```{r}

```

### 1.4. Monthly temperatures in 1924 for all locations

```{r}

```


## Exercise 2: Detective Work for Athletes Data

### 2.1. Find the incorrect entry and obtain the correct value

```{r}

```

### 2.2. Best three variables for separating sprinter and 400m groups

```{r}

```
